---
title: "COVID19_steps"
author: "SridharRamachandran"
date: "2024-07-21"
output:
  pdf_document: 
    toc: true
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include=FALSE}
library(tidyverse)
library(lubridate)
```

## Introduction

The files for the COVID 19 Data Report project was sourced from https://github.com/CSSEGISandData/COVID-19. The data source is in CSV format. We are going to use data in this folder https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series. This folder contains daily time series summary tables, including confirmed, deaths. 

## Importing Data
I will start by reading in the data from the four main csv files

```{r get_jhu_data}
# get current data from four files
url_in = "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names = c("time_series_covid19_confirmed_global.csv", 
               "time_series_covid19_deaths_global.csv", 
               "time_series_covid19_confirmed_US.csv", 
               "time_series_covid19_deaths_US.csv")
urls = str_c(url_in, file_names)

```

## Reading the dataset
```{r import_data, message=FALSE}
global_cases = read_csv(urls[1])
global_deaths = read_csv(urls[2])
us_cases = read_csv(urls[3])
us_deaths = read_csv(urls[4])

```
## Global Cases
### Tidying the data
Dates, cases and deaths need to be in their own column
Remove Latitude and Longitude as it will not be used in analysis
Rename Region and State fields

```{r tidy_global_data}
global_cases = global_cases %>%
  pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long), 
               names_to = "date", 
               values_to = "cases") %>%
  select(-c(Lat, Long))

global_deaths = global_deaths %>%
  pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long), 
               names_to = "date", 
               values_to = "deaths") %>%
  select(-c(Lat, Long))

```

### Transform the data
```{r transform_global_data}
global = global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = `Country/Region`, Province_State = `Province/State`) %>%
  mutate(date = mdy(date))

```
### Summary of data
```{r summary_global}
summary(global)
```
### Filter out the data and re-summarize
```{r filter_global}
global = global %>%
  filter(cases > 0)

summary(global)

```
## US Data
### Tidying the data
```{r tidy_us_data}
us_cases = us_cases %>%
  pivot_longer(cols = -(UID:Combined_Key), names_to = "date", values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

us_deaths = us_deaths %>%
  pivot_longer(cols = -(UID:Population), names_to = "date", values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))
```

### Transform the data
```{r join_us_data}
us = us_cases %>%
  full_join(us_deaths) 

```
### Add population data to global data
```{r global_add_pop_data}
global = global %>%
  unite("Combined_Key", c(Province_State, Country_Region), 
        sep = ",", 
        na.rm = TRUE, 
        remove = FALSE)

uid_lookup_url = "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid_lookup = read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))


global = global %>%
  left_join(uid_lookup, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)


```
## Visualizing Data
### US data
```{r us_by_state}
US_by_State = us %>%
  group_by(Province_State, Country_Region, date) %>%
  summarise(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000/Population) %>%
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population)%>%
  ungroup()

```
```{r us_by_country_region}
US_totals = US_by_State %>%
  group_by(Country_Region, date) %>%
  summarise(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000/Population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_mill, Population)%>%
  ungroup()

```
```{r plot1}
US_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID-19 in US",y = NULL)



```
```{r plot2}

state = "Washington"
US_by_State %>%
  filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID-19 in ", state),y = NULL)

```
## Data Analysis

Is the number of new cases basically flat? Basically there are no new cases because things have leveled off. Well, in order to analyze this, we are going to add to our data. What we're going to do is we will transform our data again, and what we will do is we will add new variables. We will call new cases to be the cases minus the lag of cases and new death. So we will add these two columns to our datasets.

```{r add_variables_plot3}
US_by_State = US_by_State %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

US_totals = US_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

US_totals %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_Cases")) +
  geom_point(aes(color = "new_Cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID-19 in US", y = NULL)

state = "Washington"
US_by_State %>%
  filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID-19 in ", state),y = NULL)

```
```{r plot4}
US_state_totals = US_by_State %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases), Population = max(Population),
            cases_per_thou = 1000 * cases/Population, deaths_per_thou = 1000 * deaths/Population) %>%
  filter(cases > 0, Population > 0)



```

## Modeling data
Linear Model. deaths per 1,000 is considered as a function of the cases per 1,000 

### Create Model

```{r model_data}
mod = lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)
summary(mod)

```
### Prediction based on Linear Model

```{r pred}
x_grid = seq(100, 500)
new_df = tibble(cases_per_thou = x_grid)
us_tot_w_pred = US_state_totals %>%
  mutate(pred = predict(mod))

```

### Plot the graph with prediction

```{r plot5}
us_tot_w_pred %>%
  ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red")

```

